### 2023-11-09 manuscript/manuscript.Rmd
Update the abstract to include the results from the standards mix

```{r assignment-summary}
# tar_load(assignment_results)
# tar_load(assignment_matches)
# tar_load(spike_urine_standards_MF_assignment_possible)
# tar_load(spiked_urine_correct_assignments)
# 
# total_assigned <- round(sum(assignment_results$n_assignments) / 
#                           sum(assignment_results$n_features) * 100,1)
# total_mfs <- sum(assignment_results$n_MFs)
# total_kegg <- sum(assignment_matches$KEGG_compound_matches)
# 
# spiked_urine_possible <- nrow(spike_urine_standards_MF_assignment_possible)
# spiked_urine_correct_percent <- spiked_urine_correct_assignments$MF %>% 
#   unique() %>% 
#   length() %>% 
#   {. / spiked_urine_possible * 100} %>% 
#   round(digits = 1)
```

We present a data-driven ensemble approach for the ‘first pass’ automated assignment of molecular formulas to mass spectra acquired from high-resolution electrospray ionisation mass spectrometry (ESI-MS). This is primarily intended for metabolomics experiment data using non-targeted metabolome fingerprinting and LC-MS profiling.
The approach incorporates the use of *m/z* correlation networks, molecular formula generation and selection as well as graphically based component plausibility selection to assign molecular formulas to *m/z* features.
The R package *assignments* was developed as a user-friendly implementation of the approach and is freely available from https://github.com/aberHRML/assignments.
The approach was tested in human urine samples spiked with known chemical standards and a data set comparing leaf tissue samples of different ecotypes of the model grass species *Brachypodium distachyon*. Across the three test matrices, `r #total_assigned`% of the m/z features were assigned a molecular formula giving a total of `r #total_mfs` molecular formulas and matching to `r #total_kegg` compounds in the KEGG compound database. Of the `r #spiked_urine_possible` spiked chemical standards that it was possible to assign in the human urine samples, the approach was able to assign m/z features to `r #spiked_urine_correct_percent`% of these.

@@

```{r assignment-evaluations}
tar_load(standards_parameters_molecular_formula_assignment)
tar_load(standards_assignment_outcomes_non_iso)
tar_load(spiked_urine_assignment_outcomes_non_iso)
tar_load(standards_compound_outcomes)

@@

standards_n_alternative <- standards_assignment_outcomes_non_iso %>% 
  dplyr::filter(
    dplyr::if_all(
      correlations:assigned,
      ~.x == TRUE
    ),
    matching_adduct == FALSE
  ) %>% 
  nrow()

compounds_percent_assigned <- standards_compound_outcomes %>% 
  dplyr::filter(
    outcome %in% c(
      "MFs matched to m/z features",
      "Total MFs assigned"
    )
  ) %>% 
  tidyr::gather(
    sample,n,
    -outcome
  ) %>% 
  tidyr::spread(
    outcome,n
  ) %>% 
  dplyr::group_by(sample) %>% 
  dplyr::summarise(
    percent = {`Total MFs assigned` / `MFs matched to m/z features` * 100} %>% 
      signif(digits = 3)
  ) %>% 
  dplyr::pull(percent) %>% 
  purrr::set_names(c(
    'spiked_urine',
    'standards'
  )) %>% 
  as.list()
```

@@

```{r assignment-results}
tar_read(assignment_results) %>%
  dplyr::full_join(
    tar_read(assignment_matches) %>% 
      dplyr::mutate(
        matrix = matrix %>% 
          replace(
            matrix == 'standards',
            'standards-mix'
          )
      ), 
    by = c("matrix", "technique")) %>% 
  mutate(
    matrix = paste(matrix,technique)
  ) %>%
  mutate(
    n_assignments = glue::glue(
      '{n_assignments} ({signif(n_assignments/n_features * 100,digits = 3)}\\%)'),
    across(where(is.numeric),~prettyNum(.x,big.mark = ','))) %>%
  select(-technique) %>% 
  dplyr::rename(
    `\\textit{m/z} features` = n_features,
    correlations = n_correlations,
    relationships = n_relationships,
    `adduct and isotopic step assignments` = n_ai_assignments,
    `transformation step assignments` = n_t_assignments,
    `assigned \\textit{m/z}` = n_assignments,
    `molecular formulas (MF)` = n_MFs,
    `KEGG compound matches` = KEGG_compound_matches,
    `KEGG MF matches` = KEGG_MF_matches,
    `matched chemical standards` = matched_standards) %>% 
  tidyr::gather(Variable,Value,-matrix) %>% 
  mutate(
    Value = replace(Value,
                    Value == 'NA',
                    'N/A')) %>% 
  tidyr::spread(matrix,Value) %>% 
  slice(c(1,4,9,2,10,3,8,6,5)) %>%
  dplyr::select(
    Variable,
    `standards-mix FIE-HRMS`,
    `spiked-urine FIE-HRMS`,
    `brachy FIE-HRMS`,
    `brachy C18-UHPLC-HRMS`
  ) %>% 
  set_colnames(
    stringr::str_split_fixed(colnames(.),
                             ' ',2)[,2]) %>% 
  knitr::kable(
    format = 'latex',
    booktabs = TRUE,
    linesep = '',
    escape = FALSE,
    align = c(
      rep('l',1),
      rep('r',4)
    ),
    caption = 'A summary of the molecular formula assignment results for the example sample matrices.') %>%
  kable_styling(
    font_size = 10,
    latex_options = "HOLD_position"
  ) %>%
  add_header_above(c(
    " ",
    "Standards mix" = 1,
    "Human urine" = 1,
    "\\\\textit{B. distachyon} leaf" = 2),
    escape = FALSE)
```

>>>>

```{r assignment-summary}
tar_load(standards_compound_outcomes)
tar_load(assignment_results)
tar_load(assignment_matches)

compounds_possible <- standards_compound_outcomes %>% 
  dplyr::filter(
    outcome == 'MFs matched to m/z features'
  ) %>% 
  dplyr::pull(
    standards
  )

compounds_percent_assigned <- standards_compound_outcomes %>% 
  dplyr::filter(
    outcome %in% c(
      "MFs matched to m/z features",
      "Total MFs assigned"
    )
  ) %>% 
  tidyr::gather(
    sample,n,
    -outcome
  ) %>% 
  tidyr::spread(
    outcome,n
  ) %>% 
  dplyr::group_by(sample) %>% 
  dplyr::summarise(
    percent = {`Total MFs assigned` / `MFs matched to m/z features` * 100} %>% 
      signif(digits = 3)
  ) %>% 
  dplyr::pull(percent) %>% 
  purrr::set_names(c(
    'spiked_urine',
    'standards'
  )) %>% 
  as.list()

average_percent_assigned <- assignment_results %>% 
  dplyr::mutate(
    percent_assigned = n_assignments / n_features * 100
  ) %>% 
  dplyr::pull(percent_assigned) %>% 
  mean() %>% 
  signif(digits = 3)

average_kegg_compounds <- assignment_matches %>% 
  dplyr::pull(KEGG_compound_matches) %>% 
  mean(na.rm = TRUE) %>% 
  round()
```

We present a data-driven ensemble approach for the ‘first pass’ automated assignment of molecular formulas to mass spectra acquired from high-resolution electrospray ionisation mass spectrometry (ESI-MS). This is primarily intended for metabolomics experiment data using non-targeted metabolome fingerprinting and LC-MS profiling.
The approach incorporates the use of *m/z* correlation networks, molecular formula generation and selection as well as graphically based component plausibility selection to assign molecular formulas to *m/z* features.
The R package *assignments* was developed as a user-friendly implementation of the approach and is freely available from https://github.com/aberHRML/assignments.
The approach was tested in a mix of known chemical standards, human urine samples spiked the chemical standards mix and a data set comparing leaf tissue samples of different ecotypes of the model grass species *Brachypodium distachyon*. 
Of the `r compounds_possible` chemical standards that had theoretical ionisation products matching *m/z* features in the standards mix, the approach assigned `r compounds_percent_assigned$standards`% of these.
Across the example matrices, an average of `r average_percent_assigned`% of the *m/z* features were assigned a molecular formula matching an average of `r average_kegg_compounds` compounds in the KEGG compound database.

@@

```{r assignment-evaluations}
tar_load(standards_parameters_molecular_formula_assignment)
tar_load(standards_assignment_outcomes_non_iso)
tar_load(spiked_urine_assignment_outcomes_non_iso)

@@

standards_n_alternative <- standards_assignment_outcomes_non_iso %>% 
  dplyr::filter(
    dplyr::if_all(
      correlations:assigned,
      ~.x == TRUE
    ),
    matching_adduct == FALSE
  ) %>% 
  nrow()
```

@@

```{r assignment-results}
assignment_results %>%
  dplyr::full_join(
    assignment_matches %>% 
      dplyr::mutate(
        matrix = matrix %>% 
          replace(
            matrix == 'standards',
            'standards-mix'
          )
      ), 
    by = c("matrix", "technique")) %>% 
  mutate(
    matrix = paste(matrix,technique)
  ) %>%
  mutate(
    n_assignments = glue::glue(
      '{n_assignments} ({signif(n_assignments/n_features * 100,digits = 3)}\\%)'),
    across(where(is.numeric),~prettyNum(.x,big.mark = ','))) %>%
  select(-technique) %>% 
  dplyr::rename(
    `\\textit{m/z} features` = n_features,
    correlations = n_correlations,
    relationships = n_relationships,
    `adduct and isotopic step assignments` = n_ai_assignments,
    `transformation step assignments` = n_t_assignments,
    `assigned \\textit{m/z}` = n_assignments,
    `molecular formulas (MF)` = n_MFs,
    `KEGG compound matches` = KEGG_compound_matches,
    `KEGG MF matches` = KEGG_MF_matches,
    `matched chemical standards` = matched_standards) %>% 
  tidyr::gather(Variable,Value,-matrix) %>% 
  mutate(
    Value = replace(Value,
                    Value == 'NA',
                    'N/A')) %>% 
  tidyr::spread(matrix,Value) %>% 
  slice(c(1,4,9,2,10,3,8,6,5)) %>%
  dplyr::select(
    Variable,
    `standards-mix FIE-HRMS`,
    `spiked-urine FIE-HRMS`,
    `brachy FIE-HRMS`,
    `brachy C18-UHPLC-HRMS`
  ) %>% 
  set_colnames(
    stringr::str_split_fixed(colnames(.),
                             ' ',2)[,2]) %>% 
  knitr::kable(
    format = 'latex',
    booktabs = TRUE,
    linesep = '',
    escape = FALSE,
    align = c(
      rep('l',1),
      rep('r',4)
    ),
    caption = 'A summary of the molecular formula assignment results for the example sample matrices.') %>%
  kable_styling(
    font_size = 10,
    latex_options = "HOLD_position"
  ) %>%
  add_header_above(c(
    " ",
    "Standards mix" = 1,
    "Human urine" = 1,
    "\\\\textit{B. distachyon} leaf" = 2),
    escape = FALSE)
```
